%!TEX encoding = UTF-8 Unicode

%% Thesis cover pages
%%
%% Based on the example.tex file provided by Tomás Oliveira e Silva
%%
%% Modified by João Paulo Barraca <jpbarraca@ua.pt>
%% Modified by Filipe Manco <filipe.manco@ua.pt>

% Remove 'coverdepname' parameter to remove department name from cover pages
% Add 'final' parameter to remove "DOCUMENTO PROVISÓRIO" from cover page
\documentclass[11pt,twoside,a4paper]{report}

%%% This file is a template for the cover of your
%%%  thesis.
%%% Lines like this one, starting with %%% , are
%%%  instructions. Search them to know where and
%%%  how to edit this file. After editing the
%%%  document you can delete them.
%%% Items marked with OPTIONAL can be removed.
%%% Most of the lines you will have to change
%%%  already have examples to help you understand.
%%% The complete list of options and commands can
%%%  be found in uaThesisTemplate.sty


%%% Select your area of study and department from
%%%  the following lists and replace them in the
%%%  \usepackage declaration ([area,department]).
%%%
%%% | area        |   | department |
%%% ---------------   --------------
%%% | arts        |   | dao        |
%%% | sciences    |   | dbio       |
%%% | education   |   | dcspt      |
%%% | economy     |   | deca       |
%%% | engineering |   | degeit     |
%%% | humanities  |   | de         |
%%% | health      |   | deti       |
%%% |             |   | demac      |
%%% |             |   | decivil    |
%%% |             |   | dem        |
%%% |             |   | dfis        |
%%% |             |   | dlc        |
%%% |             |   | dmat       |
%%% |             |   | dq         |
%%% |             |   | dcm        |
%%% |             |   | dgeo        |
%%% ---------------   --------------
%% Import the UA thesis template
\usepackage[engineering,deti]{uaThesisTemplate}


%% Optional packages for utf8 encoding and T1 fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}

%% For prettier pages
\usepackage{lmodern}

%% We will use english and portuguese
\usepackage[english,portuguese]{babel}

\usepackage{xspace}% used by \sigla

%%% This package is only used to create a dummy
%%%  text for the abstract. You can remove the
%%%  following line in the final document.
%\usepackage{blindtext}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%       DEFINE THESIS COVER FIELDS
%%                  START
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%% Edit thesis year
%% Thesis Year
\ThesisYear{\the\year}


%%% Uncomment one of the following options:
%\DocumentTypeThesis
%\DocumentTypeThesisProposal
\DocumentTypeDissertation

%%% Uncomment if you are a student of a MAP Doctoral program 
%%% http://www.map.edu.pt/
%%% | area                | identifier |
%%% | Biology of Plants   | bioplant   |
%%% | Computer Science    | i          |
%%% | Physics             | fis        |
%%% | Telecommunications  | tele       |

%\MAPProgram{i}

%%% Edit author's name
%% Author's full name
\Author{JOÃO ANTÓNIO \\ASSIS REIS}

%%% Edit thesis title in Portuguese
%% Thesis Mandatory Title in Portuguese
\TitlePT{UM SISTEMA CONVERSACIONAL DE PESQUISA SOBRE BASES DE DADOS MÉDICAS}

%%% OPTIONAL: you can write your thesis title in
%%%  other languages. You can repeat the following
%%%  line as many times as you want, or remove it
%%%  if you only want the portuguese title.
%% Optional Titles in additional languages.
\TitleEN{A CONVERSATIONAL QUERY BUILDER ON MEDICAL DATABASES}

%%% OPTIONAL: add a quotation if you like.
%% Optional Quotation
%% \Quote{\selectlanguage{english}The greatest challenge to any thinker is stating
%% the problem in a way that will allow a solution}{Bertrand Russell}

%%% Edit your supervisor's information.
%%%
%%% Use this command for male supervisor
%%% \Supervisor{name}{category}
%%%
%%% Or this command for femmale supervisor
%%% \Supervisora{name}{category}
%%%
%%% name   : Supervisor's name
%%% categoy: One of the following list
%%%
%%% | category                  |
%%% -----------------------------
%%% | catedrático/a             |
%%% | catedrático/a convidado/a |
%%% | associado/a c/ agregação  |
%%% | associado/a               |
%%% | auxiliar                  |
%%% | auxiliar convidado        |
%%% -----------------------------
%% Supervisor information

%%% OPTIONAL: you can add a second supervisor.
%%%
%%% Use this command for male supervisor
%%% \SecondSupervisor{name}{category}
%%% 
%%% Or this command for femmale supervisor
%%% \SecondSupervisora{name}{category}
%%%  {[preposition: (at, in, no, da)] department}
%%%  {[preposition: (at, in, do, na)] school}
%%%
%%% name      : See above.
%%% category  : See above.
%%% department: Can be a department, faculty or
%%%             other; Choose the preposition
%%%             accordingly.
%%% school    : University/Scholl name; Choose the
%%%             preposition accordingly.
%% Optional SecondSuperviser information
\Supervisor{João Rafael Almeida}{auxiliar}

\SecondSupervisor{José Luís Oliveira}{catedrático}{do Departamento de Electrónica, Telecomunicações e Informática}{da Universidade de Aveiro}

%%% Edit your degree information.
%%%
%%% \Degree{title}{name}
%%%
%%% title: Degree title (Mestre or Doutor)
%%% name : Degree name
%% Degree title and name
\Degree{Mestre}{Engenharia Informática}

%%% OPTIONAL: edit grant support texts where
%%%  appropriate. Maximum two.
%%%
%%% \GrantText{grant1}{grant2}
%% Optional Grant support texts
% \GrantText{Texto Apoio financeiro do POCTI no âmbito do III Quadro Comunitário de Apoio.}
%           {Texto Apoio financeiro da FCT e do FSE no âmbito do III Quadro Comunitário de Apoio.}

%%% OPTIONAL: Add a dedication text.
%% Optional Dedication text
% \Dedication{Dedico este trabalho à minha esposa e filho pelo incansável apoio.}

%%% OPTIONAL: Add an acknowledgement text
%% Optional Acknowledgement text
\Acknowledgements{Em primeiro lugar, expresso minha profunda gratidão aos meus orientadores, especialmente ao Professor Auxiliar João Rafael Almeida. O seu apoio foi além da orientação do projeto; guiou-me e desafiou-me ao longo de todo este processo, motivando-me a superar e enfrentar novos desafios. Agradeço também ao colaborador desta dissertação, Tiago Almeida, que sempre esteve disponível para me aconselhar e guiar no melhor caminho.

A todos os meus amigos, estou imensamente grato pelo constante companheirismo e pela partilha da vida académica ao longo dos últimos anos. À Inês, um agradecimento especial pela paciência e pelo suporte incondicional.

Por fim, dedico este trabalho à minha família que sempre me apoiou e tornou possível a minha jornada académica até aqui.
}

%% commented


%%% Edit the keywords and abstract in Portuguese
%%% Don't forget to remove the \blindtext[3]
%%%  declaration.
%%% Don't remove the \selectlanguage statement or you will end with incorrect hyphenation
%% Mandatory Keywords and Abstract in Portuguese
\Abstract{Palavras Chave}{Portal EHDEN, Descoberta de dados biomédicos, Partilha de dados, OMOP CDM, Sistema conversacional de pesquisa, Bases de dados médicas, Recuperação de Informação, OHDSI, Definição de Coortes}
         {Resumo}{\selectlanguage{portuguese}Na investigação médica, os estudos multicêntricos tornaram-se viáveis devido às estratégias atuais que abordam questões de interoperabilidade entre bases de dados. No entanto, à medida que essas estratégias foram amplamente adotadas, novos desafios na descoberta de dados emergiram. Inicialmente, esses desafios foram enfrentados com catálogos de bases de dados e painéis gráficos para análise e comparação de dados, mas essas soluções tornaram-se menos eficazes à medida que as comunidades de pesquisa se expandiram. Esta dissertação aborda esta questão desenvolvendo um motor de busca conversacional para ajudar os investigadores médicos a encontrar de forma eficiente as bases de dados mais adequadas para os seus estudos. Além disso, pretende melhorar este motor para auxiliar os investigadores na definição dos seus protocolos de estudo. A metodologia inclui a implementação da técnica BM25 para a recuperação ideal de bases de dados e a utilização de um grande modelo de linguagem (LLM) para tarefas de geração de texto e processamento de linguagem natural (PLN). Os resultados demonstram que o sistema conversacional de pesquisa melhora significativamente a eficiência na descoberta de bases de dados, reduzindo o tempo e o esforço necessários para os investigadores identificarem bases de dados apropriadas. Além disso, a integração da definição de coortes simplifica o processo complexo de definição de coortes, melhorando assim o fluxo geral da investigação.}

%%% OPTIONAL: you can write the keywords and
%%%  abstract in other languages. You can repeat
%%%  the following lines as many times as you want
%%%  or remove them if you only want portuguese.
%%% Don't remove the \selectlanguage statement or you will end with incorrect hyphenation.
%%% You can modify the language to any other supported by Latex
%% Optional Keywords and Abstract in additional languages.
\Abstract{Keywords}{EHDEN Portal, Biomedical data discovery, Data sharing, OMOP CDM, Conversational query builder, Medical databases, Information Retrieval, OHDSI, Cohort Definition}
         {Abstract}{In medical research, multicentre studies have become feasible due to current strategies addressing interoperability issues between databases. However, as these strategies have been widely adopted, new challenges in data discovery have emerged. Initially tackled with database catalogues and graphical dashboards for data analysis and comparison, these solutions become less effective as research communities expand. This dissertation addresses this issue by developing a conversational search engine to help medical researchers efficiently find the most suitable databases for their studies. Additionally, it aims to enhance this engine to assist researchers in defining their study protocols. The methodology includes implementing the BM25 technique for optimal database retrieval and utilizing a large language model (LLM) for text generation and natural language processing (NLP) tasks. Results demonstrate that the conversational query builder significantly improves the efficiency of database discovery, reducing the time and effort required for researchers to identify appropriate databases. Furthermore, integrating cohort definition simplifies the complex process of defining cohorts, thereby enhancing the overall research workflow.}

%%% Edit the Jury President information.
%% First Member will be the President of the Jury

% \JuriMember{Prof. Doutor João Antunes da Silva}{professor associado da Faculdade de Engenharia da Universidade do Porto}

%%% Repeat the above declaration for additional
%%%  members of the jury as many times as necessary.
%% Additional members of the Jury

% \JuriMember{Prof. Doutor João Antunes da Silva}{professor associado da Faculdade de Engenharia da Universidade do Porto}



\IA{I acknowledge using Grammarly (Grammarly, https://www.grammarly.com/) and ChatGPT-3.5 (Open AI, https://chat.openai.com/) to improve some sentences, namely grammatical structure, punctuation, and vocabulary.}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%       DEFINE THESIS COVER FIELDS
%%                  END
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


\begin{document}

%%% By default we consider portuguese to be the
%%%  main language (the last one in the usepackage
%%%  declaration). If you're writing your thesis
%%%  in english uncomment the following line.
%%% Set the proper language or you will end with incorrect hyphenation!
\selectlanguage{english}


\MakeCover
\end{document}
