\chapter{Conclusions}
\label{chapter:Conclusions}

% sumário dos findings, contribuições chave, bm25 fucniona ou não, chat é bom nisto ou não, future work, no trabalho deu origem a três artigos

%% REFERIR OS OBJETIVOS E DIZER QUE FORAM CUMPRIDOS

% 1) Study of state-of-the-art: i) methodologies to build a conversational user interface, ii) procedures to retrieve the databases of most interest, and iii) explore the definition of a study protocol;
% 2) Developed a chat-like search engine to help discover the best databases for a study;
% 3) Enhance the engine to collect additional information to provide a query as an outcome. 

The proposed system addresses some real and meaningful challenges in the medical research field, essentially the complexity of finding the best databases in a vast catalogue and defining a study protocol in the ATLAS platform. It led to the creation of three significant scientific papers that contributed to the advancement of the field.

Regarding implementation, it is possible to extract some conclusions about the choices, strategy, and areas. About BM25, it shows promising results, but creating an annotated dataset may enable us to identify how to improve these results or other powerful {\ir} techniques. To add the {\llm} into the system, FlowiseAI shows to be an excellent tool for the development of {\llm}-based applications. However, it is limited due to this project's complexity, specifically the query builder phase. So, the option to integrate the {\llm} is a Python-based backend developed for this project system.

The introduction of this conversational search tool addresses the significant challenge of navigating a growing repository of medical databases. This system enables users to engage in natural language dialogue, significantly reducing the time and complexity traditionally associated with identifying suitable databases for research studies. Moreover, the system reflects a deep understanding of the needs within the medical research community for reliable, up-to-date, and accessible data.

Also, introducing a conversational query builder is innovative due to the limited existence of a query build that works in a conversational manner. This feature of the system not only saves time for medical researchers but also saves the researcher from the complexity of defining a cohort or a concept set in the current tool, ATLAS.

In conclusion, the three objectives, mentioned in the Chapter \ref{chapter:Introduction}, are accomplised with success. The extensive literature review on conversational interfaces, {\nlp}, and {\ir} was crucial to develop a chat-like search engine to help discover the best databases for a study and enhance this engine to collect additional information to provide a query as an outcome. The successful implementation of this objectives represents a significant advancement in the field of data retrieval, discovery and analysis. It not only simplifies the process of identifying the most relevant databases, but also helps in the definition of study protocols process, empowering researchers, regardless of their prior experience with database search tools and query builders.

However, some work remains to be done in the future, namely to enhance the query builder feature in order to provide the full cohort definition. Due to the complexity of a cohort definition, the current system only attempts the concept set definition and the observation window, which are parts of the cohort. More work is needed to enhance this system, such as improving the user interface, enhancing the {\ir} methods to produce better results, and creating synthetic labeled data to test and validate the {\ir} methods.

This system has the potential to be very useful for medical researchers and play a significant role in driving technological progress.
 
